---
title: "Introduction to BLASTr"
vignette: >
  %\VignetteIndexEntry{Introduction to BLASTr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    eval: false
---

```{r}
#| label: setup
#| eval: true
library(BLASTr)
```

`BLASTr` runs BLAST+ searches from R, tailored for taxonomic
classification of Amplicon Sequence Variants (ASVs) from metabarcoding
and metagenomic experiments. All chunks below are not evaluated when the
package is built (they require network access and external tools), but
can be run interactively.

## Installing command-line dependencies

`BLASTr` manages its command-line dependencies (BLAST+, Entrez Direct,
and seqkit) in dedicated conda environments through the `condathis`
package.
The first call to any function that needs a tool will install it
automatically; you can also install everything up front:

```{r}
#| label: install-deps
install_dependencies()
```

Pinned tool versions are installed for reproducibility. To upgrade the
environments later, run `install_dependencies(force = TRUE)`.

## Building a BLAST database

Any FASTA file with unique headers can be turned into a BLAST+ database:

```{r}
#| label: make-db
fasta_path <- fs::path_package(
  "BLASTr", "extdata", "minimal_db_blast",
  ext = "fasta"
)
taxid_map <- fs::path_package(
  "BLASTr", "extdata", "minimal_db_blast",
  ext = "txt"
)
db_path <- fs::path_temp("minimal_db_blast")

make_blast_db(
  fasta_path = fasta_path,
  db_path = db_path,
  db_type = "nucl",
  taxid_map = taxid_map
)
```

The optional `taxid_map` file (one `<SequenceId> <TaxonomyId>` pair per
line) makes BLAST results include subject Tax IDs (`staxid`), which can
then be resolved to full lineages with `parallel_get_tax()`.

## Searching sequences in parallel

`parallel_blast()` deduplicates the query sequences, groups them into
multi-record FASTA batches, and dispatches the batches to parallel
workers (`mirai` daemons). Failed batches are salvaged per-sequence and
retried.

```{r}
#| label: parallel-blast
asvs <- readLines(
  fs::path_package("BLASTr", "extdata", "asvs_test", ext = "txt")
)

blast_res <- parallel_blast(
  query_seqs = asvs,
  db_path = db_path,
  total_cores = 2,
  num_alignments = 4
)

blast_res
```

The result has one row per unique query sequence, with the top
`num_alignments` hits spread into `1_`, `2_`, ... column groups
(subject header, identity, alignment coordinates, e-value, bit score,
query coverage, and subject Tax ID).

Every query's BLAST exit status and error/warning output is preserved:

```{r}
#| label: exit-codes
exit_codes(blast_res)
```

## Retrieving taxonomy

Tax IDs from the BLAST hits can be resolved into full taxonomic
lineages using NCBI Entrez Direct. Requests are batched (up to 100 Tax
IDs per request):

```{r}
#| label: get-tax
tax_ids <- unique(blast_res$`1_staxid`)

tax_tbl <- parallel_get_tax(
  organisms_taxIDs = tax_ids,
  total_cores = 1
)

tax_tbl
```

You can also search NCBI Taxonomy by organism name:

```{r}
#| label: get-tax-name
get_tax_by_name("Danio rerio")
```

Note that NCBI enforces API rate limits (3 requests per second without
an API key, 10 with one). Entrez Direct honors the `NCBI_API_KEY`
environment variable automatically.

## Performance notes

- **Batching** is the main speed lever: each BLAST+ process loads the
  database index once and searches many queries. Control it with
  `chunk_size` (default: automatic, targeting ~4 chunks per core).
- **`total_cores`** sets how many BLAST+ processes run concurrently;
  `num_threads` sets threads per process. Effective CPU usage is
  `total_cores * num_threads`.
- If the [mori](https://cran.r-project.org/package=mori) package is
  installed, the deduplicated query vector is shared with workers
  through OS shared memory (zero-copy), which matters for very large
  query sets.
- If your session already has `mirai::daemons()` set, `BLASTr` uses the
  existing pool instead of creating a new one.
